Longitudinal Study of Software Environments Produced by Dockerfiles from Research Artifacts: Initial Design

### 2025-04-04 src/main.tex
[text] submittable version

The entire scientific community has been struck by the ``Reproducibility Crisis''~\cite{baker500ScientistsLift2016}, and Computer Science does not make exception~\cite{collberg_repeatability_2015}.
In order to improve the reproducibility of articles in Computer Science, the community set up \adae\ processes~\cite{kidwell2016badges} where authors would \emph{voluntary} submit the artifact required to reproduced the main results from their article.
Then, artifact reviewers would try to reproduce those main results given the artifact from the authors.
Based on the result of the reproduction attempt, the article is then awarded \emph{badges} to promote its reproducibility aspect~\cite{acm-badges}.
The link to the artifact can be found in the published version of the article, sometimes with an appendix giving more information about the artifact (\eg\ download instructions, hardware requirements, workflow)~\cite{paperswithcode, kang2023papers, hong2013software}.

One difficult aspect of creating a truly reproducible artifact is to properly control the software environment: packages and libraries versions.
Several works already shed light on the possible variability of results given a badly captured software environment \cite{mytkowicz_producing_nodate, sokolowski2024impact}.
To tackle this issue, the conference \adae\ committees have been advising authors to provide containers or virtual machines to share their artifacts in order to ``freeze'' their software environment~\cite{eurosys25, aetips, csartifacts}.
%However, authors generate their artifact right before the artifact submission~\cite{guilloteau2024longevity, guilloteau2024frustrations}.
In the case of usual virtualization techniques to generate containers or virtual machines, authors have to write the ``recipe'' made of a list of commands to execute from a ``base image'' (for example, \df\ for Docker containers).
This approach has several limitations.
First, these base images can vary, either because of bad practices from the users (\eg\ using the \texttt{latest} version of the images) or from the administrators of the image (\eg\ pushing a different image to a previously defined tag).
These base images can also be removed by the administrators making all the recipes unbuildable~\cite{nvidia_cuda_lifetime}.
Another limitation of this approach is the heavy dependency on non-reproducible tools such as classical package managers (\eg\ \texttt{apt}, \texttt{dpkg}, \texttt{yum}).
Hence, rebuilding a recipe of a container or a virtual machine can yield different results from the authors' image, and thus threatens the reproducibility of the results.
One solution would be for the authors to store the built container itself (\ie\ the tarball) onto some long term storage for example Zenodo~\cite{zenodo}.
But this come with the downside of requiring more storage space than just storing the recipe (binary vs. text file), and thus has some \textbf{severe scalability and sustainability issues in terms of longevity}~\cite{monroe2023preservation, guilloteau2024longevity}.
Moreover, storing the built binary image exhibits limitations when needed to \emph{inspect} the actual content of the image, and when a modification of the software environment capture by this image is needed~\cite{mercier2018considering}.
Having the opportunity to extend previous scientific work is a crucial aspect in any scientific field.

@@

With \ecg, a \df\ from an artifact is represented as a Nickel configuration file~\cite{nickel}.
Nickel is a configuration language with functional paradigm.
One of Nickel's features is the possibility to \emph{verify} that a configuration file is valid with respect to a \emph{contract}.
The Nickel file representing the artifact is then \emph{verified} against the contract~\cite{ecg_contract} and translated to a JSON file given as input to \ecg.
The contract is versioned to take into account the potential future changes to the descriptions and the contract.
Listing~\ref{lst:nickel} shows an example of Nickel representation.

@@

\subsubsection{Hash of the Artifact}

The articles give the link to the artifact.
This artifact can be shared in several ways, but not all ways have the same guarantees in terms of longevity and ``stability'' of the link.
Hence, we capture the result of the download at the given link (success or error).
In case of a successful download, we also compute the cryptographic hash of the result of the download to capture any variation in the source of the artifact which could later explain variations in the resulting software environment.

@@

In the case of a \df s failing to build, we capture the error (missing image, failing command, etc.).
In fact, we are only able to capture the \emph{first error} of the build even if there might be several errors in the building process.
The Docker error is extracted by \texttt{grep}ing the building process log.

@@

The software environment can be composed of several packages, which can have been installed by different means.
The most common way is through package managers (\eg\ \texttt{apt}, \texttt{dpkg}, \texttt{yum}).
But there are also installed packages for programming languages such as with \texttt{pip} for Python.
Another way to install package is to download the sources and build them locally.
\ecg\ is able to capture these different ways to install packages.
Hence, for each package, we extract the package name, the version, and the tool used to install the package.

As of for now, \ecg\ is able to query information from popular package managers (\texttt{dkpg} (and thus \texttt{apt}), \texttt{pacman}, \texttt{pip}, \texttt{conda}) and to deal with situations where packages are install by source (\texttt{package\_managers} in Listing~\ref{lst:nickel}).
In the case where there is a \texttt{git clone} command in the \df\ (\texttt{git\_packages} in Listing~\ref{lst:nickel}), \ecg\ will extract the commit of the local clone of the repository after the successful build of the container.
This allows us to trace the version of the package.
In the case of a simple download of the sources of a package (\texttt{misc\_packages} in Listing~\ref{lst:nickel}), for example with \texttt{wget} or \texttt{curl}, there is no immediate way to log the version of the package that is being downloaded.
The URL of the source code might contain a version number, but this URL might in the future point to a different version of the source code.
Hence, \ecg\ will download \emph{locally} the source code pointed by the URL and compute the cryptographic hash of the result to generate a ``version''.

@@

\begin{figure*}
  \centering
  %\fbox{
  \includegraphics[width=1\textwidth,trim={1.40cm 1.65cm 1.65cm 1.65cm},clip]{./figs/workflow.pdf}
  %}
  \caption{
    Workflow of \ecg.
    Each description of an artifact is verified with the Nickel contract and then converted in a JSON representation.
    This JSON representation is then read by \ecg\ to
    \textit{(i)} download the artifact,
    \textit{(ii)} compute the hash of its content,
    \textit{(iii)} build the container from the \df, and
    \textit{(iv)} extract the software environment information from the built container.
    \ecg\ outputs files containing the information about the artifact and its \df\ .
  }
  \label{fig:workflow}
\end{figure*}

The workflow of \ecg\ is depicted in Figure~\ref{fig:workflow}.
% We make sure to use an empty local Docker cache before building the containers.
The workflow itself \cite{ecg_workflow} is ruled by a Snakemake~\cite{koster2012snakemake} description and its software environment \cite{ecg_nix} is managed with Nix~\cite{dolstra_nix_2004}.

\subsection{Contributions}

This framework leaves a lot of room for contributions on the writing of the Nickel descriptions files.
The community could contribute by submitting pull requests on the open source repository containing all the Nickel descriptions.
\todo{more}

@@

\begin{enumerate}
\item For each conference with an artifact evaluation process, we look at all the artifacts and consider only the artifacts using a \df\ to share their software environment.
\item From the artifact and the \df, we write the associated Nickel file (Listing~\ref{lst:nickel}): where is stored the artifact? how to extract the \df\ out of the artifact? which package managers are used?
\item Once all the Nickel descriptions for the artifacts of a conference have been written, we use \ecg\ to build the download, build, and extract the information out of each of them.
%\item We run \ecg\ the closer to the start of each month as possible.
\item The local Docker cache of these machine is cleaned before the building of each container.
\item We will execute \ecg\ on each artifact every month, as close as possible to the start of the month, for a full year (\ie\ 13 months).
\end{enumerate}

@@

% \todo{}
% 
% \begin{itemize}
% \item we convert the dockerfiles to nickel description
% \item periodic rebuild of the dockerfiles
% \item g5k
% \item empty cache
% \item max 10 mins
% \item contributions
% \end{itemize}

% \newpage
\section{Preliminary Results}\label{sec:results}

%This section presents the information we would like to extract from the data gathered.

% And there is no ``warning message'' that the environment is different.

We wrote the descriptions in Nickel (see Section~\ref{sec:nickel}) for the five artifacts of the EuroPar'24 conference that used a Docker container to share their software environments, and used \ecg\ to build the associated \df s once every month for seven months.
% "2024-10-01" "2024-11-04" "2024-12-02" "2025-01-02" "2025-02-05" "2025-03-12"
The initial build was October 1$^{st}$ 2024, then the following builds were on November 4$^{th}$ 2024, December 2$^{nd}$ 2024, January 2$^{nd}$ 2025, February 5$^{th}$ 2025, March 12$^{th}$ 2025, and April 1$^{st}$ 2025.
The deadline to submit the artifacts was May 15$^{th}$ 2024\footnote{\url{https://2024.euro-par.org/calls/artifacts/}}.
%\todo{when was the artifact submission, and publicaiton of the papers for europar?}
The data collected is available on Zenodo~\cite{dataset_europar24}.

The following sections follow the dimensions captured by \ecg\ (see Section~\ref{sec:ecg:capture}).

@@

While this initial study only considered five \df s, it revealed that \textbf{the software environments produced by the considered \df s variate \emph{every month}} which highlights the lack of reproducibility of such practices.
The question of the significance of these results on such a small sample will need to be answered in a larger scale study.

\textbf{
  A practice to deal with the lack of reproducibility of the \df s is for the authors to store the built container onto some long-time storage such as Zenodo.
  However, this practice raises serious questions in terms of scalability and sustainability in terms of storage.
  The lack of reproducibility of \df s should not be hidden with long-term storage: the practices from the conferences, journals, and the authors need to change.
}

>>>>

The entire scientific community has been struck by the ``Reproducibility Crisis''~\cite{baker500ScientistsLift2016}, and Computer Science does not make exception~\cite{collberg_repeatability_2015}.
In order to improve the reproducibility of articles in Computer Science, the community set up, in conferences and journals, \adae\ processes~\cite{kidwell2016badges} where authors would \emph{voluntary} submit the artifact required to reproduced the main results from their article.
Then, artifact reviewers would try to reproduce those main results given the artifact from the authors.
Based on the result of the reproduction attempt, the article is then awarded \emph{badges} to promote its reproducibility aspect~\cite{acm-badges}.
The link to the artifact can be found in the published version of the article, sometimes with an appendix giving more information about the artifact (\eg\ download instructions, hardware requirements, workflow)~\cite{paperswithcode, kang2023papers, hong2013software}.

One difficult aspect of creating a truly reproducible artifact is to properly control the software environment: packages and libraries versions.
Several works already shed light on the possible variability of results given a badly captured software environment~\cite{mytkowicz_producing_nodate, sokolowski2024impact}.
To tackle this issue, the conference \adae\ committees have been advising authors to provide containers or virtual machines to share their artifacts in order to ``freeze'' their software environment~\cite{eurosys25, aetips, csartifacts}.
%However, authors generate their artifact right before the artifact submission~\cite{guilloteau2024longevity, guilloteau2024frustrations}.
In the case of usual virtualization techniques to generate containers or virtual machines, authors have to write the ``recipe'' (for example, \df\ for Docker containers) made of a list of commands to execute from a ``base image'' (the \texttt{FROM} instruction in \df s).
This approach has several limitations.
First, these base images can vary, either because of bad practices from the users (\eg\ using the \texttt{latest} version of the images) or from the administrators of the image (\eg\ pushing a different image to a previously defined tag).
These base images can also be removed by the administrators making all the recipes unbuildable~\cite{nvidia_cuda_lifetime}.
Another limitation of this approach is the heavy dependency on non-reproducible tools such as classical package managers (\eg\ \texttt{apt}, \texttt{dpkg}, \texttt{yum}).
Hence, rebuilding a recipe of a container or a virtual machine can yield different results from the authors' image, and thus threatens the reproducibility of the results.
One solution would be for the authors to store, and then provide, the built container itself (\ie\ the tarball) onto some long term storage (for example Zenodo~\cite{zenodo}).
But this come with the downside of requiring more storage space than just storing the recipe (binary vs. text file), and thus has some \textbf{severe scalability and sustainability issues in terms of longevity}~\cite{monroe2023preservation, guilloteau2024longevity}.
Moreover, storing the built binary image exhibits limitations when needed to \emph{inspect} the actual content of the image, and when a modification of the software environment capture by this image is needed~\cite{mercier2018considering}.
Having the opportunity to extend previous scientific work is a crucial aspect in any scientific field.

@@

With \ecg, a \df\ from an artifact is represented as a Nickel configuration file~\cite{nickel}.
Nickel is a configuration language with functional paradigm.
One of Nickel's features is the possibility to \emph{verify} that a configuration file is valid with respect to a \emph{contract}.
The Nickel file representing the artifact is then \emph{verified} against the contract~\cite{ecg_contract} and translated to a JSON file given as input to \ecg.
Listing~\ref{lst:nickel} shows an example of Nickel representation.
The contract is versioned (line 2 of Listing~\ref{lst:nickel}) to take into account the potential future changes to the descriptions and the contract.

@@

\subsubsection{Cryptographic Hash of the Artifact}

Artifacts can be shared in several ways by the authors, but not all ways have the same guarantees in terms of longevity and ``stability'' of the link.
Hence, \ecg\ captures the result of the download at the given link (success or error).
In case of a successful download, \ecg\ computes the cryptographic hash of the result of the download to capture any variation in the source of the artifact which could later explain variations in the resulting software environment.

@@

In the case of a \df\ failing to build, \ecg\ capture the error returned by the building process (\eg\ missing image, failing command).
In fact, \ecg\ is only able to capture the \emph{first error} of the build even if there might be several errors in the building process.
The Docker error is extracted by \texttt{grep}ing the building process log.

@@

The software environment can be composed of several packages, which can have been installed by different means.
The most common way is through package managers (\eg\ \texttt{apt}, \texttt{dpkg}, \texttt{yum}).
But there are also installed packages for programming languages such as with \texttt{pip} for Python.
Another way to install package is to download the sources and build them locally.
\ecg\ is able to capture these different ways to install packages.
Hence, for each package, \ecg\ extracts the package name, the version, and the tool used to install the package.

As of for now, \ecg\ is able to query information from popular package managers (\texttt{dkpg} (and thus \texttt{apt}), \texttt{pacman}, \texttt{pip}, \texttt{conda}) and to deal with situations where packages are install by source (\texttt{package\_managers} in Listing~\ref{lst:nickel}), or the case of packages installed in virtual Python environments.
In the case where there is a \texttt{git clone} command in the \df\ (\texttt{git\_packages} in Listing~\ref{lst:nickel}), \ecg\ will extract the commit of the local clone of the repository after the successful build of the container, and use this commit as a ``version''.
In the case of a download of the sources of a package (\texttt{misc\_packages} in Listing~\ref{lst:nickel}), for example with \texttt{wget} or \texttt{curl}, there is no immediate way to log the version of the package that is being downloaded.
The URL of the source code might contain a version number, but this URL might in the future point to a different version of the source code.
Hence, \ecg\ will download \emph{locally} (outside of the container) the source code pointed by the URL and compute the cryptographic hash of the result to generate a ``version''.

@@

\begin{figure*}
  \centering
  \includegraphics[width=1\textwidth,trim={1.37cm 1.65cm 1.65cm 1.65cm},clip]{./figs/workflow.pdf}
  \caption{
    Workflow of \ecg.
    Each description of an artifact is verified with the Nickel contract and then converted in a JSON representation.
    This JSON representation is then read by \ecg\ to
    download the artifact,
    compute the hash of its content,
    build the container from the \df, and
    extract the software environment information from the built container.
    \ecg\ outputs files containing the information about the artifact and its \df\ .
  }
  \label{fig:workflow}
\end{figure*}

The workflow of \ecg\ is depicted in Figure~\ref{fig:workflow}.
The workflow itself~\cite{ecg_workflow} is ruled by a Snakemake~\cite{koster2012snakemake} description and its software environment~\cite{ecg_nix} is managed with Nix~\cite{dolstra_nix_2004}.

% \subsection{Contributions}
% 
% This framework leaves a lot of room for contributions on the writing of the Nickel descriptions files.
% The community could contribute by submitting pull requests on the open source repository containing all the Nickel descriptions.
% \todo{more}

@@

\begin{enumerate}
\item For each conference with an artifact evaluation process, we look at all the artifacts and consider only the artifacts using a \df\ to share their software environment.
\item From the artifact and the \df, we write the associated Nickel file (Listing~\ref{lst:nickel}): where is stored the artifact? how to extract the \df\ out of the artifact? which package managers are used?
\item Once all the Nickel descriptions for the artifacts of a conference have been written, we use \ecg\ to build the download, build, and extract the information out of each of them.
\item The local Docker cache of these machine is cleaned before the building of each container.
\item We will execute \ecg\ on each artifact every month, as close as possible to the start of the month, for a full year (\ie\ 13 months).
\end{enumerate}

@@

\section{Preliminary Results}\label{sec:results}

We wrote the descriptions in Nickel (see Section~\ref{sec:nickel}) for the five artifacts of the EuroPar'24 conference that used a Docker container to share their software environments, and used \ecg\ to build their associated \df s once every month for seven months.
% "2024-10-01" "2024-11-04" "2024-12-02" "2025-01-02" "2025-02-05" "2025-03-12" "2025-04-01"
The initial build was October 1$^{st}$ 2024, then the following builds were on November 4$^{th}$ 2024, December 2$^{nd}$ 2024, January 2$^{nd}$ 2025, February 5$^{th}$ 2025, March 12$^{th}$ 2025, and April 1$^{st}$ 2025.
The deadline for the authors to submit their artifacts at EuroPar'24 was on May 15$^{th}$ 2024\footnote{\url{https://2024.euro-par.org/calls/artifacts/}}.
Hence our initial build took place almost 5 months after the artifacts submission.

The following sections follow the dimensions captured by \ecg\ (see Section~\ref{sec:ecg:capture}).
The data collected, and presented in this section, is available on Zenodo~\cite{dataset_europar24}.

@@

While this initial study only considered five \df s, it revealed that \textbf{the software environments produced by the considered \df s variate \emph{every month}} which highlights the lack of reproducibility of such tools and methods to share software environments.
The question of the results' significance on such a small sample will need to be answered in a larger scale study.

\textbf{
  A practice to deal with the lack of reproducibility of the \df s is for the authors to store the built container onto some long-time storage such as Zenodo.
  However, this practice raises serious questions in terms of scalability and sustainability in terms of storage~\cite{monroe2023preservation}.
  The lack of reproducibility of \df s should not be hidden with long-term storage: the practices of the authors and of the \adae\ need to be re-evaluated.
}

The community might want to adopt solutions \emph{tailored for reproducibility} of software environment (\ie\ Nix~\cite{dolstra_nix_2004} and Guix~\cite{courtes_functional_2013}), where authors can share a textual reprenstation of their software environment which could be precisely reproduced, and with longevity guarentees~\cite{courtes2024source}.
Moreover, Nix and Guix allow for the precise introduction of variation in the software environment, which is a crucial feature (not present in container technologies) in order to build upon previous artifact for future publications~\cite{mercier2018considering}.
